OptiFit: a fast method for fitting amplicon sequences to existing OTUs

### 2021-06-17 paper/paper.Rmd
Create figure for analysis workflow

```{r knitr_settings, eval=TRUE, echo=FALSE, cache=FALSE}
schtools::set_knitr_opts()
```

```{r deps, message = FALSE}
library(here)
library(schtools)
library(tidyverse)
load(here('results', 'stats.RData'))
```
<!--
- **"AND, BUT, THEREFORE" structure. start paragraph with question, end with why we should care. transitions to move the story along.**
- From Pat: "briefly looking through the Discussion and Intro, one point that we may have forgotten is that a benefit of our approach is that it is much easier to customize to a specific region. The greengenes reference OTUs are based on full length sequences. This causes problems when considering shorter (e.g. V4) sequences since reference OTUs may be more similar  than the full length and even identical to each other in the subregion. Because we can easily spin up a subregion specific set of reference OTUs from a public database or the reference fraction this isn't a problem. This is described in one of my earlier papers looking at open/closed reference clustering and was part of the reason that the order of the database was important."
-->

@@

While _de novo_ clustering produces high quality OTUs, researchers may prefer to
perform reference clustering to a public database because reference-based
methods produce consistent OTUs and are generally faster than _de novo_ methods.
In closed reference mode, sequences that cannot be assigned to reference OTUs
are thrown out, so that the final clustering contains only OTUs that exist in
the reference.
To test how OptiFit performs for this purpose, we fit each dataset to three
databases of reference OTUs: the Greengenes database, the SILVA non-redundant
database, and the Ribosomal Database Project (RDP) [@desantis_greengenes_2006;
@quast_silva_2013; @cole_ribosomal_2014].
Reference OTUs for each database were created by performing _de novo_ clustering
with OptiClust at a distance threshold of 3%.
The _de novo_ MCC scores were `r dn_dbs_mcc` for `r names(dn_dbs_mcc)`, 
respectively.
Fitting sequences to Greengenes and SILVA in closed reference mode performed
similarly, with median MCC scores of `r closed_fit_gg_mcc` and 
`r closed_fit_silva_mcc` respectively, while the median MCC 
dropped to `r closed_fit_rdp_mcc` when fitting to RDP.
For comparison, clustering datasets _de novo_ with
OptiClust produced an average MCC score of `r opticlust_mcc`.
This gap in OTU quality mostly disappeared when clustering in open reference
mode, which produced median MCCs of `r open_fit_gg_mcc` with greengenes, 
`r open_fit_silva_mcc` with SILVA, and `r open_fit_rdp_mcc` with RDP.
Thus, open reference OptiFit produced OTUs of very similar quality as _de novo_
clustering, and closed reference OptiFit followed closely behind as long as a
suitable reference database was chosen.

>>>>

```{r knitr_settings, eval=TRUE, echo=FALSE, cache=FALSE}
schtools::set_knitr_opts()
knitr::opts_chunk$set(fig.path = here::here('figures/'))
```

```{r deps, message = FALSE}
library(cowplot)
library(here)
library(knitr)
library(schtools)
library(tidyverse)
load(here('results', 'stats.RData'))
```
<!--
- **"AND, BUT, THEREFORE" structure. start paragraph with question, end with why we should care. transitions to move the story along.**
- From Pat: "briefly looking through the Discussion and Intro, one point that we may have forgotten is that a benefit of our approach is that it is much easier to customize to a specific region. The greengenes reference OTUs are based on full length sequences. This causes problems when considering shorter (e.g. V4) sequences since reference OTUs may be more similar  than the full length and even identical to each other in the subregion. Because we can easily spin up a subregion specific set of reference OTUs from a public database or the reference fraction this isn't a problem. This is described in one of my earlier papers looking at open/closed reference clustering and was part of the reason that the order of the database was important."
-->

@@

While _de novo_ clustering produces high quality OTUs, researchers may prefer to
perform reference clustering to a public database because reference-based
methods produce consistent OTUs and are generally faster than _de novo_ methods.
In closed reference mode, sequences that cannot be assigned to reference OTUs
are thrown out, so that the final clustering contains only OTUs that exist in
the reference.
To test how OptiFit performs for this purpose, we fit each dataset to three
databases of reference OTUs: the Greengenes database, the SILVA non-redundant
database, and the Ribosomal Database Project (RDP) [@desantis_greengenes_2006;
@quast_silva_2013; @cole_ribosomal_2014].
Reference OTUs for each database were created by performing _de novo_ clustering
with OptiClust at a distance threshold of 3% (see Figure \@ref(fig:workflow)).
The _de novo_ MCC scores were `r dn_dbs_mcc` for `r names(dn_dbs_mcc)`, 
respectively.
Fitting sequences to Greengenes and SILVA in closed reference mode performed
similarly, with median MCC scores of `r closed_fit_gg_mcc` and 
`r closed_fit_silva_mcc` respectively, while the median MCC 
dropped to `r closed_fit_rdp_mcc` when fitting to RDP.
For comparison, clustering datasets _de novo_ with
OptiClust produced an average MCC score of `r opticlust_mcc`.
This gap in OTU quality mostly disappeared when clustering in open reference
mode, which produced median MCCs of `r open_fit_gg_mcc` with greengenes, 
`r open_fit_silva_mcc` with SILVA, and `r open_fit_rdp_mcc` with RDP.
Thus, open reference OptiFit produced OTUs of very similar quality as _de novo_
clustering, and closed reference OptiFit followed closely behind as long as a
suitable reference database was chosen.

```{r workflow, dpi=300, fig.width=6, fig.cap="The Analysis Workflow"}
tiff::readTIFF(here('figures', 'workflow.tiff')) %>% 
  grid::grid.raster() 
```
